Experience Report: Haskell in Computational Biology

### 2012-07-01 mrfy_experience_report.tex
fixed the nonlocal pairing for beta strands, sort of
this was a lot harder than I thought, and I couldn't get an equeation

To~account for beta strands, computational biologists add
a term to Viterbi's equations, and they build
richer models of protein structure.
%%  The~richer model recognizes that only certain pairs of amino acids are
%%  able to bond to each other.
%%  The presence of beta strands changes both the natu
%%   the probability that a given
%%  query sequence fit
%%  A~mutation in a paired beta strand may break the pairing, in which
%%   case the mutated protein will not fold properly and is unlikely to
%%   survive natural selection.
%%  But a mutation in a paired beta strand \emph{can} survive if it is
%%   accompanied by a \emph{compensatory} mutation in the strand with which it is
%%   paired. 
When column~$j$ of an alignment is part of a beta strand and is paired
with another column  $\pairedwith j$,
the probability of finding amino acid~$x_i$ in column~$j$ 
depends on~$x_{\pairedwith i}$.
The~relevant term in
Viterbi's equations 
$V_{j}^{\prime M}(i)$ becomes
$$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log P(x_{i} \mid x_{\pairedwith i}).$$
Columns~$j$ and $\pairedwith j$ can be as close together as a few
columns or as far apart as a few hundreds of columns.
Because $W_j^M(i)$~depends not only on $x_i$ and its local neighbors but
also on the nonlocal $x_{\pairedwith j}$, 
dynamic programming cannot compute the maximum likelihood quickly
\cite{Menke:2010ti,Daniels:2012}.

>>>>

To~account for beta strands, computational biologists add
a term to Viterbi's equations, and they build
richer models of protein structure.
%%  The~richer model recognizes that only certain pairs of amino acids are
%%  able to bond to each other.
%%  The presence of beta strands changes both the natu
%%   the probability that a given
%%  query sequence fit
%%  A~mutation in a paired beta strand may break the pairing, in which
%%   case the mutated protein will not fold properly and is unlikely to
%%   survive natural selection.
%%  But a mutation in a paired beta strand \emph{can} survive if it is
%%   accompanied by a \emph{compensatory} mutation in the strand with which it is
%%   paired. 
When column~$j$ of an alignment is part of a beta strand and is paired
with another column  $\pairedwith j$,
the probability of finding amino acid~$x_i$ in column~$j$ 
depends on whatever amino acid~$x'$  is in column~${\pairedwith i}$.
If~$x'$ is located in position~$i'$ in the query sequence, Viterbi's
equations must be altered so that, for example
$V_{j}^{\prime M}(i)$ depends not only on
$V_{j-1}^{\prime M}(i-1)$ but also on
$V_{\pairedwith j}^{\prime M}(i')$.\remark{I can't get an equation I
believe in. ---NR}
The distance between $j$~and~$\pairedwith j$ can be as small as a few
columns or as large as a few hundreds of columns.
Because $V_j^{\prime M}(i)$~depends not only on nearby values but also on
$V_{\pairedwith j}^{\prime M}(i')$,
dynamic programming cannot compute the maximum likelihood quickly
\cite{Menke:2010ti,Daniels:2012}.
